Stochastic Approximation Monte Carlo with Dynamic Update Factor (SAD)

### 2018-08-20 papers/histogram/sad.tex
more text on convergence

{\color{red}
It is for this reason that we propose a way to determine the update factor
``dynamically'' while still satisfying the criteria for convergence.
For the update factor to scale well across both discrete and continuous systems,
it must include the number of states $N_S$. We define the average entropy as
$\bar S$ and the time since the simulation last found a new energy state $t_L$.
We then can write the following expression for the \emph{proposed} update factor.
\begin{align}
\gamma = N_S \frac{\bar S + \nicefrac{t}{t_L}}{N_{S}^2 + \nicefrac{t^2}{t_L}}
\end{align}
We now consider the following limiting cases:
\subsubsection{$t \propto N_S \approx t_L$ and $ t << N_S$}
In this limiting case, we find that the update factor reduces to
\begin{align}
\gamma \approx \frac{\bar S}{N_S} \approx \frac{\bar S}{t}
\end{align}
Because $t \propto N_S$, the update factor scales as 1/t which is consistent
with the requirement for SAMC convergence.
\subsubsection{$t >> N_{S}^2$ and $ t\approx t_L$}
In this limiting case, we find that the update factor reduces to
\begin{align}
\gamma \approx \frac{N_S \bar S t_L}{t^2} \approx \frac{N_S \bar S}{t}
\end{align}
\subsubsection{$t >> t_L \bar S$}
In this limiting case, we find that the update factor reduces to
\begin{align}
\gamma \approx \frac{N_S}{t}
\end{align}
We now integrate (sum) the update factor from the last time an energy state was found
to the final simulation time to ensure that convergence is guaranteed.
\begin{align}
\int_{t_L}^{t_f} \gamma dt \approx t_L N_S \left[ \int_{t_L}^{t_f}
\frac{\bar S -1}{t^2} dt + \frac{1}{t_L}\int_{t_L}^{t_f} \frac{1}{t} dt \right]
\end{align}
From this we obtain
\begin{align}
\int_{t_L}^{t_f} \gamma dt \approx t_L N_S \left[\left(\bar S -1\right)\left(\frac{1}{t_L}-\frac{1}{t_f}\right) + \frac{1}{t_L} \ln\left(\frac{t_f}{t_L}\right) \right]
\end{align}
Simplifying, we see that as $t_f \rightarrow \infty$ then $\gamma \rightarrow \infty$ satisfying
the requirement for SAMC convergence.
\begin{align}
\gamma \approx \left(1-\frac{t_L}{t_f}\right)N_S\left(\bar S -1\right) + N_S \ln\left(\frac{t_f}{t_L}\right)
\end{align}
}

@@

@@

>>>>

@@

\subsubsection{Convergence}
The difficulty in using the SAMC method lies in identifying an appropriate
value for $t_0$.  A value that is either too high or too low will result
in slow convergence to the true entropy.  It is instructive to consider
separately values of $t_0$ that are too low or too high.

We can place a rigorous \emph{lower} bound $t_{\min}$ on the number of
moves required to find the true energy by considering the total change
that needs to be made to the entropy.
\begin{align}
  \Delta S_{\text{tot}} \equiv \sum_E S(E) - S_{\min}
\end{align}
The minimum number of moves that could converge the entropy is
the number of moves that will enable the above total entropy change,
which we can approximate using an integral:
\begin{align}
   \Delta S_{\text{tot}} &= \sum_{t=0}^{t_{\min}} \gamma_t \\
  &= \int_0^{t_{\min}} \gamma(t)dt \\
  &= t_0 + \int_{t_0}^{t_{\min}} \frac{t_0}{t}dt
  \\
  &= t_0\left(1 + \ln\left(\frac{t_{\min}}{t_0}\right)\right)
\end{align}
Solving for ${t_{\min}}$ we find that
\begin{align}
  {t_{\min}} &= t_0 e^{\frac{\Delta S_{\text{tot}}}{t_0} - 1}
\end{align}
which means that the minimum time to convergence grows exponentially
as $t_0$ is made smaller.  You \emph{seriously} don't want to underestimate
$t_0$!

One might reasonably choose to err on the high side when selecting a $t_0$.
The rate of convergence is harder to estimate when $t_0$ is large, but
in general $\gamma_t$ itself forms a lower bound on the accuracy with which
the entropy may be known, with an unknown prefactor which is roughly
the coherence time of the Monte Carlo simulation.  Since $\gamma_t$ is
given by $t_0/t$, the time to converge to a given accuracy is increased
in proportion to the ratio by which we overestimate $t_0$.  Thus, while
it is exponentially painful to underestimate $t_0$, overestimating by
several orders of magnitude is also not acceptable.  We should note
that these extreme limiting cases do not preclude the possibility that there
is a wide range of $t_0$ values that lead to an acceptable convergence
rate.

@@

\subsection{Convergence}
{\color{red}
It is for this reason that we propose a way to determine the update factor
``dynamically'' while still satisfying the criteria for convergence.
For the update factor to scale well across both discrete and continuous systems,
it must include the number of states $N_S$. We define the average entropy as
$\bar S$ and the time since the simulation last found a new energy state $t_L$.
We then can write the following expression for the \emph{proposed} update factor.
\begin{align}
\gamma = N_S \frac{\bar S + \nicefrac{t}{t_L}}{N_{S}^2 + \nicefrac{t^2}{t_L}}
\end{align}
We now consider the following limiting cases:
\subsubsection{$t \propto N_S \approx t_L$ and $ t << N_S$}
In this limiting case, we find that the update factor reduces to
\begin{align}
\gamma \approx \frac{\bar S}{N_S} \approx \frac{\bar S}{t}
\end{align}
Because $t \propto N_S$, the update factor scales as 1/t which is consistent
with the requirement for SAMC convergence.
\subsubsection{$t >> N_{S}^2$ and $ t\approx t_L$}
In this limiting case, we find that the update factor reduces to
\begin{align}
\gamma \approx \frac{N_S \bar S t_L}{t^2} \approx \frac{N_S \bar S}{t}
\end{align}
\subsubsection{$t >> t_L \bar S$}
In this limiting case, we find that the update factor reduces to
\begin{align}
\gamma \approx \frac{N_S}{t}
\end{align}
We now integrate (sum) the update factor from the last time an energy state was found
to the final simulation time to ensure that convergence is guaranteed.
\begin{align}
\int_{t_L}^{t_f} \gamma dt \approx t_L N_S \left[ \int_{t_L}^{t_f}
\frac{\bar S -1}{t^2} dt + \frac{1}{t_L}\int_{t_L}^{t_f} \frac{1}{t} dt \right]
\end{align}
From this we obtain
\begin{align}
\int_{t_L}^{t_f} \gamma dt \approx t_L N_S \left[\left(\bar S -1\right)\left(\frac{1}{t_L}-\frac{1}{t_f}\right) + \frac{1}{t_L} \ln\left(\frac{t_f}{t_L}\right) \right]
\end{align}
Simplifying, we see that as $t_f \rightarrow \infty$ then $\gamma \rightarrow \infty$ satisfying
the requirement for SAMC convergence.
\begin{align}
\gamma \approx \left(1-\frac{t_L}{t_f}\right)N_S\left(\bar S -1\right) + N_S \ln\left(\frac{t_f}{t_L}\right)
\end{align}
}
